Preventing runtime errors of Redis at compile time

### 2015-11-19 paper.tex
brushed up

In this paper, we demonstrates how to prevent those runtime errors at compile
time, allowing users to write more reliable database queries without runtime
overhead, by exploiting type-level programming techniques such as indexed
monad, type-level literals and closed type families in Haskell.

@@

\paragraph{Categories and Subject Descriptors}
\paragraph{Keywords}
type-level programming; Haskell; database query language; Redis;
\end{abstract}

@@

Redis is an open source, in-memory data structure store, used as
database, cache and message broker. Each value is associated with a binary-safe
string key to identify and manipulate with. Redis supports many different
kind of values, such as strings, hashes, lists and sets.

@@

The previous example in Redis's client would look something like this with
Hedis in Haskell.
\begin{minted}[xleftmargin=1em]{haskell}
program :: Redis (Either Reply [ByteString])
program = do
    sadd "some-set" ["a", "b"]
    sadd "another-set" ["a", "b", "c"]
    sinter ["some-set", "another-set"]
\end{minted}

Function \mintinline{haskell}{sadd} takes a key and a list of values as
arguments, and returns an \mintinline{haskell}{Integer}, wrapped in
\mintinline{haskell}{Either} to indicate possible failures in the context
\mintinline{haskell}{Redis}\footnotemark of command execution.

@@

\footnotetext{Hedis provides another kind of context, \mintinline{haskell}{RedisTx}, for \emph{transactions}, united with \mintinline{haskell}{Redis} under the class of \mintinline{haskell}{RedisCtx}. We use \mintinline{haskell}{Redis} only here for brevity. }

@@

Although everything in Redis is essentially a binary string, they are treated
differently. Redis supports many different kind of data structures, such as
strings, hashes, lists, etc. These values have different datatypes, and most
commands only work with certain types of them, much like how C language treats
a piece of data.

\paragraph{Problem 1} Consider the following example. \mintinline{text}{SET} is
a command that only works on strings. The key \mintinline{text}{some-string} is
now related with a string. If we treat it as a set and add an element to it with
\mintinline{text}{SADD}, a runtime error arises.

@@

Haskell makes it easy to build and use Domain Specific Languages (DSLs),
and Hedis can be regarded as one of them. What makes Hedis special is that,
it has variable bindings (between keys and values), but with very
little or no semantic checking, neither dynamically nor statically.

@@

\begin{itemize}[noitemsep]
\item We make Hedis statically type-checked, without runtime overhead.
\item We demonstrates how to model variable bindings of an embedded DSL with
language extensions like type-level literals and data kinds.
\item We provide (yet another) an example of encoding effects and
 constraints of an action in types, with indexed monad\cite{indexedmonad}
  and other language extensions such as closed
  type-families\cite{closedtypefamilies} and constraints
  kinds\cite{constraintskinds}.
\item Edis, a package we built for programmers. This package helps programmers
 to write more reliable Redis codes, and also makes Redis polymorphic by
 automatically converting back and forth from values of any types and boring
 ByteStrings.
\end{itemize}

@@

\footnotetext{\mintinline{haskell}{TypeRep} here represents term-level
representations of datatypes, available in \mintinline{haskell}{Data.Typeable}}

@@

Luckily, with recently added GHC extension \emph{data kinds}. With data kinds
we can define our own kinds just like we can with datatypes, and every suitable
datatype will be promoted to be a kind, and its value constructors promoted
to be type constructors, automatically.

@@

Haskell sugars lists \mintinline{haskell}{[1, 2, 3]} and
tuples \mintinline{text}{(1, 'a')} with brackets and parentheses.
Promoted lists and tuples are also sugared with single quote prefixed. For
example \mintinline{text}{'}\mintinline{haskell}{[Int, Char]},
\mintinline{text}{'}\mintinline{haskell}{(Int, Char)}.

@@

\begin{minted}[xleftmargin=1em]{haskell}
"this is a type-level string literal" :: Symbol
\end{minted}

Nonetheless, it's still useful to have a term-level value that links with a
Symbol, when we want to retrieve type-level information at runtime (but not the
 other way around!).

@@

They are defined with \emph{type synonym}, since they are \emph{types}
, not \emph{terms}. If we ask \mintinline{text}{GHCi} what is the kind of
\mintinline{haskell}{Dict1}, we get \mintinline{haskell}{Dict1 :: [ (Symbol, *) ]}

The star \mintinline{haskell}{*} is a kind, it stands for any kind of types,
such as \mintinline{haskell}{Int}, \mintinline{haskell}{Char} or even a symbol
\mintinline{haskell}{"symbol"}, while \mintinline{haskell}{Symbol} stands only
for all symbols.

@@

Class \mintinline{haskell}{IMonad} comes with two operations:
\mintinline{haskell}{unit} for identities and \mintinline{haskell}{bind} for
compositions, as in Monad.

We define a new datatype \mintinline{haskell}{Edis}, which is basically just
\mintinline{haskell}{Redis} indexed with more information in types. And we make
 it an instance of \mintinline{haskell}{IMonad}.

@@

To execute a \mintinline{haskell}{Edis} program, simply apply it to
 \mintinline{haskell}{unEdis} to get an ordinary
 \mintinline{haskell}{Hedis} program, with type information erased.

@@

\footnotetext{\mintinline{haskell}{ping} from Hedis is qualified with
\mintinline{haskell}{Hedis} to prevent function name clashing}

@@

Dictionary \mintinline{haskell}{xs} in the type remain unaffected after the
action, because \mintinline{haskell}{ping} does not affect any key-value
bindings. To encode other commands that alters key-value bindings, we need
 type-level functions to annotate those effects on the dictionary.

@@

The first instance of \mintinline{haskell}{And} could be subsumed
under the more general instance \mintinline{haskell}{And a b}.
But the closedness allows these instances to be resolved in order, just like
how cases are resolved in term-level functions. Also notice that how much
\mintinline{haskell}{And} resembles to it's term-level sibling.

@@

But \mintinline{haskell}{Get }
\mintinline{text}{'}\mintinline{haskell}{[ }
\mintinline{text}{'}\mintinline{haskell}{("A", Int) ] "B"} would get stuck.
That's because \mintinline{haskell}{Get} is a partial function on types,
 and these types are computed at compile-time. It wouldn't make
much sense for type checkers to crash and throw a "Non-exhaustive" error or
be non-terminating.

@@

    Get ’[]             s = Nothing
    Get (’(s, x) ’: xs) s = Just x
    Get (’(t, x) ’: xs) s = Get xs s
\end{minted}

Other dictionary-related functions are defined in a similar way.

@@

Since Haskell has a \emph{phase distinction}\cite{phasedistinction}, types are
erased before runtime. It's impossible to obtain information directly from
types, we can only do this indirectly, with \emph{singleton types}
\cite{singletons}.

@@

In the type of \mintinline{haskell}{del}, the type variable
\mintinline{haskell}{s} would be a \mintinline{haskell}{Symbol} that is decided
 by the argument of type \mintinline{haskell}{Proxy s}.
To use \mintinline{haskell}{del}, we would have to apply it with a clumsy
 term-level proxy like this:

@@

Commands such as \mintinline{text}{INCR}, \mintinline{text}{INCRBYFLOAT}, are
 not only container-specific, there are also some requirements on what types of
 value they could operate on.

We handle this by mapping Redis's strings of integers and floats to Haskell's
\mintinline{text}{Integer} and \mintinline{text}{Double}.

@@

The semantics of \mintinline{text}{LLEN} defined above is actually not
complete. \mintinline{text}{LLEN} also accepts keys that do not exist and
 returns 0. We will need \emph{}

@@

In other words, we require that the key to be associated with a list,
 \textbf{or} does not exist at all.

@@

But there are at least three ways to express type-level constraints
\cite{singletons}. Luckily we could express constraint disjunctions with type
 families in a modular way.

@@

With \emph{constraint kind}, a recent addition to GHC. Type constraints now has
its own kind: \mintinline{haskell}{Constraint}. That means type constraints
are not restricted to the left side of a \mintinline{haskell}{=>} anymore,
they could appear in anywhere that accepts something of kind
\mintinline{haskell}{Constraint}, and any type that has kind
\mintinline{haskell}{Constraint} can also be used as a type constraint.\footnote{See \url{https://downloads.haskell.org/~ghc/7.4.1/docs/html/users_guide/constraint-kind.html}.}

@@

Users may need to make assertions about the status of certain key-type bindings
 in a Redis program. For example, declaring the existence of a key and the type
 of its associating value. We provide these functions, which do nothing but
 fiddling with types.

@@

The following program increases the value of "A" as an integer,
and push the result of increment to list "L", and then pops it out.

@@

The syntax is pretty heavy, like the old days when there's no
do-notation\cite{history}. But if we don't need any variable bindings between
operations, we could compose these commands with a sequencing operator
\mintinline{haskell}{(>>>)}.

@@

We could even get rid of \mintinline{haskell}{(>>>)} with
 \emph{rebindable syntax}\footnote{See \url{https://downloads.haskell.org/~ghc/latest/docs/html/users_guide/syntax-extns.html#rebindable-syntax}.},
if we are so inclined.

>>>>

In this paper, we demonstrate how to prevent those runtime errors at compile
 time, thus allowing users to write more reliable database queries without
 runtime overhead, by exploiting type-level programming techniques such as
 indexed monad, type-level literals and closed type families in Haskell.

@@

\paragraph{Categories and Subject Descriptors}
\paragraph{Keywords}
Haskell; type-level programming; database query language; Redis;
\end{abstract}

@@

Redis is an open source, in-memory data structure store, used as
 database, cache and message broker. Each value is associated with a binary-safe
 string key to identify and manipulate with. Redis supports many different
 kind of values, such as strings, hashes, lists, sets, etc.

@@

The previous example in Redis's client would look something like this with
 Hedis in Haskell.

\begin{minted}[xleftmargin=1em]{haskell}
program :: Redis (Either Reply [ByteString])
program = do
    sadd "some-set" ["a", "b"]
    sadd "another-set" ["a", "b", "c"]
    sinter ["some-set", "another-set"]
\end{minted}

\mintinline{haskell}{sadd} takes a key and a list of values as arguments, and
 returns an \mintinline{haskell}{Integer}, wrapped in
 \mintinline{haskell}{Either} to indicate possible failures in the context
 \mintinline{haskell}{Redis}\footnotemark of command execution.

@@

\footnotetext{Hedis provides another kind of context, \mintinline{haskell}{RedisTx}, for \emph{transactions}, united with \mintinline{haskell}{Redis} under the class of \mintinline{haskell}{RedisCtx}. We demonstrate \mintinline{haskell}{Redis} only for brevity. }

@@

Although everything in Redis is essentially a binary string, these strings are
 treated differently. Redis supports many different kind of data structures,
 such as strings, hashes, lists, etc. These values have different datatypes,
 and most commands only work with certain types of them, much like how C
 language treats a piece of data.

\paragraph{Problem 1} Consider the following case. \mintinline{text}{SET} is
 a command that only works with strings, it associates the key
 \mintinline{text}{some-string} with a value of string. If we treat the like a
 set and add an element to it with \mintinline{text}{SADD}, a runtime error
 arises.

@@

Haskell makes it easy to build and use Domain Specific Languages (DSLs),
 and Hedis can be regarded as one of them. What makes Hedis special is that,
 it has \emph{variable bindings} (between keys and values), but with very
 little or no semantic checking, neither dynamically nor statically.

@@

\begin{itemize}[noitemsep]
\item We make Hedis statically type-checked, without runtime overhead.
\item We demonstrates how to model variable bindings of an embedded DSL with
 language extensions like type-level literals and data kinds.
\item We provide (yet another) an example of encoding effects and constraints of
 an action in types, with indexed monad\cite{indexedmonad} and other language
 extensions such as closed type-families\cite{closedtypefamilies} and
 constraints kinds\cite{constraintskinds}.
\item Edis, a package we built for programmers. This package helps programmers
 to write more reliable Redis programs, and also makes Redis polymorphic by
 automatically converting back and forth from values of arbitrary types and
 boring ByteStrings.
\end{itemize}

@@

\footnotetext{\mintinline{haskell}{TypeRep} supports term-level representations
 of datatypes, available in \mintinline{haskell}{Data.Typeable}}

@@

Luckily, with recently added GHC extension \emph{data kinds}, suitable
 datatype will be automatically promoted to be a kind, and its value
 constructors to be type constructors.

@@

Haskell sugars lists \mintinline{haskell}{[1, 2, 3]} and tuples
 \mintinline{text}{(1, 'a')} with brackets and parentheses.
 We could also express promoted lists and tuples in types like this with
 a single quote prefixed. For example:
 \mintinline{text}{'}\mintinline{haskell}{[Int, Char]},
 \mintinline{text}{'}\mintinline{haskell}{(Int, Char)}.

@@

\begin{minted}[xleftmargin=1em]{haskell}
"this is a type-level string literal" :: Symbol
\end{minted}
%
% Nonetheless, it's still useful to have a term-level value that links with a
%  Symbol, when we want to retrieve type-level information at runtime (but not the
%  other way around!).

@@

These dictionaries are defined with \emph{type synonym}, since they are
 \emph{types}, not \emph{terms}. If we ask \mintinline{text}{GHCi} what is the
 kind of \mintinline{haskell}{Dict1}, we will get
 \mintinline{haskell}{Dict1 :: [ (Symbol, *) ]}

The kind \mintinline{haskell}{*} (pronounced "star") stands for the set of all
 concrete type expressions, such as \mintinline{haskell}{Int},
 \mintinline{haskell}{Char} or even a symbol \mintinline{haskell}{"symbol"},
 while \mintinline{haskell}{Symbol} is restricted to all symbols only.

@@

Class \mintinline{haskell}{IMonad} comes with two operations:
 \mintinline{haskell}{unit} for identities and \mintinline{haskell}{bind} for
 compositions, as in monads.

We define a new datatype \mintinline{haskell}{Edis}, which is basically just
 the context \mintinline{haskell}{Redis} indexed with more information in types.
 We make it an instance of \mintinline{haskell}{IMonad}.

@@

To execute a \mintinline{haskell}{Edis} program, simply apply it to
 \mintinline{haskell}{unEdis} to get an ordinary program of type
 \mintinline{haskell}{Redis}, with type information erased.

@@

\footnotetext{\mintinline{haskell}{ping} from Hedis is qualified with
\mintinline{haskell}{Hedis} to prevent function name clashing in our code.}

@@

The dictionary \mintinline{haskell}{xs} in the type remains unaffected after the
 action, because \mintinline{haskell}{ping} does not affect any key-value
 bindings. To encode other commands that modifies key-value bindings, we need
 type-level functions to annotate those effects on the dictionary.

@@

The first instance of \mintinline{haskell}{And} could be subsumed under a more
 general instance, \mintinline{haskell}{And a b}.
But the closedness allows these instances to be resolved in order, just like
 how cases are resolved in term-level functions. Also notice that how much
 \mintinline{haskell}{And} resembles to it's term-level sibling.

@@

But \mintinline{haskell}{Get }
\mintinline{text}{'}\mintinline{haskell}{[ }
\mintinline{text}{'}\mintinline{haskell}{("A", Int) ] "B"} would get stuck.
That's because \mintinline{haskell}{Get} is a partial function on types,
 and these types are computed at compile-time. It wouldn't make
 much sense for a type checker to crash and throw a "Non-exhaustive" error or
 be non-terminating.

@@

    Get ’[]             s = Nothing
    Get (’(s, x) ’: xs) s = Just x
    Get (’(t, x) ’: xs) s = Get xs s
\end{minted}
%
Other dictionary-related functions are defined in a similar fashion.

@@

Since Haskell has a \emph{phase distinction}\cite{phasedistinction}, types are
 erased before runtime. It's impossible to obtain information directly from
 types, we can only do this indirectly, with
 \emph{singleton types}\cite{singletons}.

@@

In the type of \mintinline{haskell}{del}, the type variable
 \mintinline{haskell}{s} is a \mintinline{haskell}{Symbol} that is decided by
 the argument of type \mintinline{haskell}{Proxy s}.
 To use \mintinline{haskell}{del}, we would have to apply it with a clumsy
 term-level proxy like this:

@@

Commands such as \mintinline{text}{INCR} and \mintinline{text}{INCRBYFLOAT}, are
 not only container-specific, they also have some requirements on what types of
 value they could operate on.

We could handle this by mapping Redis's strings of integers and floats to
 Haskell's \mintinline{text}{Integer} and \mintinline{text}{Double}.

@@

The semantics of \mintinline{text}{LLEN} defined above is actually not
complete. \mintinline{text}{LLEN} also accepts keys that do not exist, and
 replies with \mintinline{text}{0}.

@@

In other words, we require that the key to be associated with a list,
 textbf{unless} it doesn't exist at all.

@@

There are at least three ways to express type-level constraints
\cite{singletons}. Luckily we could express constraint disjunctions with type
 families in a modular way.

@@

With \emph{constraint kind}, a recent addition to GHC. Type constraints now has
 its own kind: \mintinline{haskell}{Constraint}. That means type constraints
 are not restricted to the left side of a \mintinline{haskell}{=>} anymore,
 they could appear in anywhere that accepts something of kind
 \mintinline{haskell}{Constraint}, and any type that has kind
 \mintinline{haskell}{Constraint} can also be used as a type constraint.
 \footnote{See \url{https://downloads.haskell.org/~ghc/7.4.1/docs/html/users_guide/constraint-kind.html}.}

@@

Users may need to make assertions about the status of some key-type bindings in
 a Redis program. For example, declaring the existence of a key and the type
 of its associating value. We provide these functions, which do nothing but
 fiddling with types.

@@

The following program increases the value of "A" as an integer, push the result
 of the increment to list "L", and then pops it out.

@@

The syntax is pretty heavy, like the old days when there's no
 \emph{do-notation}\cite{history}. But if we don't need any variable bindings
 between operations, we could compose these commands with a sequencing operator
 \mintinline{haskell}{(>>>)}.

@@
